openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-11-02 paper/main.tex
paper: reflect new benchmarks

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2). Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification. To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{cao2019single}, containing 1.3 million single-cell gene-expression profiles. We preprocess the data using the standard single-cell analysis pipeline~\citep{kiselev2019challenges} and extract the top 50 principal components. To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size. In total, we test each implementation on 30 different data sets.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne (v0.6.2) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

% Problem with Python
In \proglang{Python}, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn}, which exhibits long runtimes when compared to other \proglang{Python} implementations. \pkg{scikit-learn}, like its \proglang{C++} counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} , is based on the Barnes-Hut approximation scheme. However, as shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer runtimes than \pkg{MulticoreTSNE}. Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points. While both libraries implement the same algorithm, \opentsne emphasizes ease of use and extensibility and is primarily written in \proglang{Python}, which inevitably introduces some runtime overhead. This makes \opentsne marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}. However, these differences disappear almost entirely when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.

@@

@@

We designed \opentsne to be accessible to a broader audience. We provide precompiled binaries for all major \proglang{Python} versions on all major platforms, making the installation process as seamless as possible. One can install \opentsne through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the \pkg{conda-forge} channel, the two most widely adopted \proglang{Python} package managers. \opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well established in the \proglang{Python} data science ecosystem. \opentsne implements multi-threaded versions of the Barnes-Hut and FIt-SNE approximation schemes, enabling it to be applied to data sets containing millions of data points. While the \proglang{Python} virtual machine inevitably introduces some performance overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}. Finally, \opentsne is extensible. Its modular design enables researchers to experiment quickly with different parameter settings and easily incorporate custom components into the software. We provide an extended list of features and comparison to other popular t-SNE implementations in Tab.~\ref{tab:features}.

>>>>

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2). Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification. To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{cao2019single}, containing 1.3 million single-cell gene-expression profiles. We preprocess the data using the standard single-cell analysis pipeline~\citep{kiselev2019challenges} and extract the top 50 principal components. To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size. In total, we test each implementation on 30 different data sets.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne (v0.6.2) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.1.2), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

% Problem with Python
In \proglang{Python}, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn}, which exhibits long runtimes when compared to other \proglang{Python} implementations. \pkg{scikit-learn}, like its \proglang{C++} counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} , is based on the Barnes-Hut approximation scheme. However, as shown in Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits somewhat longer runtimes than \pkg{MulticoreTSNE}. Newer implementations, such as \pkg{FIt-SNE} and \opentsne, implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points. While both libraries implement the same algorithm, \opentsne emphasizes ease of use and extensibility and is primarily written in \proglang{Python}. Despite this, both libraries exhibit similar runtimes, with \opentsne even being marginally faster when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.

@@

We note here that the benchmarks in Figs.~\ref{fig:benchmarks_py}~and~\ref{fig:benchmarks_lang} were run for 1,000 iterations to ensure a fair comparison between different implementations. However, it has recently been shown that, using an appropriate learning rate, we can safely reduce the number of iterations to 750~\citep{belkina2019automated}. This means that the actual runtime will be faster than reported in these benchmarks in everyday usage of \opentsne.

@@

We designed \opentsne to be accessible to a broader audience. We provide precompiled binaries for all major \proglang{Python} versions on all major platforms, making the installation process as seamless as possible. One can install \opentsne through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the \pkg{conda-forge} channel, the two most widely adopted \proglang{Python} package managers. \opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well established in the \proglang{Python} data science ecosystem. \opentsne implements multi-threaded versions of the Barnes-Hut and FIt-SNE approximation schemes, enabling it to be applied to data sets containing millions of data points. Finally, \opentsne is extensible. Its modular design enables researchers to experiment quickly with different parameter settings and easily incorporate custom components into the software. We provide an extended list of features and comparison to other popular t-SNE implementations in Tab.~\ref{tab:features}.
